---
title: "Cyclistic Rider Statistics 2022"
author: "Sharezuwan Sallehudin"
date: "2023-10-22"
output: html_document
---

### Cyclistic_Exercise_Full_Year_Analysis

#### This analysis is for case study 1 from the Google Data Analytics Certificate (Cyclistic). It's originally based on the case study "'Sophisticated, Clear, and Polished': Divvy and Data Visualization" written by Kevin Hartman (found here: <https://artscience.blog/home/divvy-dataviz-case-study>).

#### We will be using the Divvy dataset for the case study. The purpose of this script is to consolidate downloaded Divvy data into a single dataframe and then conduct simple analysis to help answer the key question: "In what ways do members and casual riders use Divvy bikes differently?"

```{r, message = FALSE}
# Load packages
library(tidyverse) # contained ggplot2
library(lubridate) # date
library(janitor)   # data cleaning
library(skimr)     # data cleaning
```

Now that necessary packages loaded, let's import our csv files.

```{r,}
# Import csv files
jan22<-read_csv("202201_divvy_tripdata.csv")
feb22 <- read_csv("202202-divvy-tripdata.csv")
mar22 <- read_csv("202203-divvy-tripdata.csv")
apr22 <- read_csv("202204-divvy-tripdata.csv")
may22 <- read_csv("202205-divvy-tripdata.csv")
jun22 <- read_csv("202206-divvy-tripdata.csv")
jul22 <- read_csv("202207-divvy-tripdata.csv")
aug22 <- read_csv("202208-divvy-tripdata.csv")
sep22 <- read_csv("202209-divvy-tripdata.csv")
sep22 <- read_csv("202209-divvy-tripdata.csv")
oct22 <- read_csv("202210-divvy-tripdata.csv")
nov22 <- read_csv("202211-divvy-tripdata.csv")
dec22 <- read_csv("202212-divvy-tripdata.csv")

# check all the files
str(jan22)
str(feb22)
str(mar22)
str(apr22)
str(may22)
str(jun22)
str(jul22)
str(aug22)
str(sep22)
str(oct22)
str(nov22)
str(dec22)
```

after checking all the csv files, now let's combined them into 1 files.

```{r}
all_trips <- bind_rows(jan22, feb22, mar22, apr22, may22, jun22, jul22, aug22, sep22, oct22, nov22, dec22)

# let's check the files
str(all_trips)
```


Now that we have our data, let's do some cleaning.
Current dataset date format is DD/MM/YYYY, let's convert it default format.

```{r}
all_trips$date <- as.Date(all_trips$started_at, format = "%d/%m/%Y")

# and let's add column that list year, month and date for each ride.
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")

# and since our day_of_week column is in numeric, let's change it to alphabet
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```

Now that we've done with the date, we have to convert the ride_length into seconds format.
Notice that started_at & ended_at column is in unproper format, let's change them first to proper datetime objects.

```{r}
all_trips$ended_at_datetime <- as.POSIXct(paste(all_trips$ended_at), format = "%d/%m/%Y %H:%M")
all_trips$started_at_datetime <- as.POSIXct(paste(all_trips$started_at), format = "%d/%m/%Y %H:%M")

# Now that we have proper datetime format, let's calculate the ride length.
all_trips$ride_length <- difftime(all_trips$ended_at_datetime, all_trips$started_at_datetime)
```

Since we have our data, notice that we have negative and 0 value ride length.
Let's do some cleaning to remove this 'bad' data.

```{r}
all_trips_v2 <- all_trips[!(all_trips$ride_length <= 0),]
```

And it's done!
Let's do descriptive analysis.

```{r}
mean(all_trips_v2$ride_length)
median(all_trips_v2$ride_length)
max(all_trips_v2$ride_length)
min(all_trips_v2$ride_length)

# Now let's compare members and casual riders
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)

# Let's see the average ride time by each day for members vs casual users
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)

# Data came out good, however the day of the week is out of order. Let's fix this.
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))

# Let's run previous code again.
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

Let's analyze ridership data by type and weekday

```{r}
all_trips_v2 %>%
  mutate(weekday = lubridate::wday(started_at_datetime, label = TRUE)) %>%
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>%
  arrange(member_casual, weekday)
```

Let's visualize the number of ride by rider type.

```{r}
all_trips_v2 %>%
  mutate(weekday = lubridate::wday(started_at_datetime, label = TRUE)) %>%
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>%
  arrange(member_casual, weekday) %>%
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) + geom_col(position = "dodge")
```

Now let's create a visualization for average duration.

```{r}
all_trips_v2 %>%
  mutate(weekday = lubridate::wday(started_at_datetime, label = TRUE)) %>%
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>%
  arrange(member_casual, weekday) %>%
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) + geom_col(position = "dodge")
```

Let's export file for further analysis.

```{r}
counts <- aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

# Conclusion
Thank you for spending your time looking at my analysis.
There is still error somewhere, please comment if I can improve more.